---
title: "Response Sheet for Modeling Soil Fluxes with NEON Data"
format: docx
editor: source
bibliography: macrosystems-tm.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Define the packages that we need to load up:
library(tidyverse)
library(lubridate)
library(bookdown)
library(knitr)
library(ggthemes)
library(plotly)
library(shiny)  # This is needed
```



**Instructions:** Read carefully and in your groups *working together* answer the numbered questions on your response paper. If you get stuck, please ask!

# Introduction


1.  Read more about NEON: [LINK](https://www.neonscience.org/about). After reviewing the information on this page, what did you wonder or notice? What would you like to know more about? **Please answer the question on the response sheet.**

2.  One site in NEON is the [San Joaquin Experimental Range](https://www.neonscience.org/field-sites/sjer) website and record three facts you find interesting or important. What scientific or ecological questions do you think could be studied here? **Please answer the question on the response sheet.**


# Fick's law

## Dependence of Fick's Law on soil water and temperature

### Understanding $g(W)$



3. Based on this information, when soils are completely dry, we would expect **more / less** diffusion compared to when soils are completely wet. **Please circle the correct answer on the response sheet.**



4. Let's assume the function $g(W)$ is **linear**. Sketch a linear equation between the two points in @fig-grid-water for $0 \leq W \leq 1$. **Sketch the function on the grid in the response sheet.** Then determine the formula for the linear function $g(W)$ on your response sheet, showing your work appropriately. **Please answer the question on the response sheet.**

```{r fig-grid-water,echo=FALSE,fig.cap="Grid for drawing $g(W)$ for different models.",fig.width=4,fig.height=4,fig.align='center',message = FALSE,warning=FALSE}


ggplot() + ylim(c(0,1)) + xlim(c(0,1)) + 
  geom_point(data = tibble(x=c(0,1),y=c(1,0)),aes(x=x,y=y),size=4,color='blue') +
  xlab("W") + ylab("g(W)") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) 


```


5. Evaluate $g(W)$ for different values of $W$, filling in @tbl-water below **on your response sheet**:

| $W$             | $0$ | $0.2$ | $0.4$ | $0.6$ | $0.8$ | $1$ |
|-----------------|-----|-------|-------|-------|-------|-----|
| $g(W)$ (linear) |     |       |       |       |       |     |
| $g(W)$ (Equation \ref{eq:moyano}) |     |       |       |       |       |     |

: Comparison of different response functions to increasing soil water content $W$. {#tbl-water}

6. Use the information in @tbl-water to add the Moyano model for $g(W)$ in @fig-grid-water. **Sketch the function on the grid in the response sheet.** What do you notice about the Moyano model in comparison to your linear model? **Please answer the question on the response sheet.**

7. Using the information in @tbl-water and @fig-grid-water, as $W$ increases, compared to the linear model the Moyano model is more / less sensitive to changes in soil water content because the linear model for $g(W)$ is greater than / less than the Moyano model at the same value of $W$. **Please circle the correct answers on the response sheet.**

8. Since $g(W)$ ultimately affects $D_{s}$, which in turn is directly proportional to $F_{s}$, interpret your previous graph to determine for what values of $W$ would cause soil flux, $F_{s}$, to increase? to decrease? **Please answer the question on the response sheet.**

### Understanding $f(T)$

9. Evaluate $f(T)$ for different values of $T$, filling in entries in @tbl-temperature below **on your response sheet**:

| $T$                  | $10$ | $15$ | $20$ | $25$ | $30$ | $35$ | $40$ | $45$ | $50$ |
|----------------------|------|------|------|------|------|------|------|------|------|
| $f(T)$ ($^{\circ}$C) |      |      |      |      |      |      |      |      |      |

: Evaluation of $f(T)$ for different soil temperatures $T$ {#tbl-temperature}

10. Next use the information in @tbl-temperature, along with other additional points, to make a plot of $f(T)$ in @fig-grid-temp. **Graph the function on your response sheet.**

```{r fig-grid-temp,echo=FALSE,fig.cap="Grid for drawing $f(T)$.",fig.width=4,fig.height=4,fig.align='center',message = FALSE,warning=FALSE}


ggplot() + ylim(c(0,2)) + xlim(c(10,50)) + 
  xlab("T (°C)") + ylab("f(T)") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) 


```

11. Based on the information in @tbl-temperature and Figure @fig-grid-temp, $f(T)$ is **increasing / decreasing** as temperature $T$ increases. **Circle the correct answer on the response sheet.**



12. Complete @tbl-temperature-day for different values of $h$, first by evaluating $T(h)$ and then by evaluating $f(T)$. **Fill in the table on the response sheet.**

| $h$                  | $0$ | $4$ | $8$ | $12$ | $16$ | $20$ | $24$ | $50$ |
|-------------------------|-----|-----|-----|------|------|------|------|------|
| $T(h)$ ($^{\circ}$C) |     |     |     |      |      |      |      |      |
| $f(T)$               |     |     |     |      |      |      |      |      |

: Evaluation of $f(T)$ for different soil temperatures $T$ {#tbl-temperature-day}

13. Finally use the information @tbl-temperature-day along with other values to sketch $f(T)$ over the course of the day in @fig-grid-temp-day-hour. **Sketch your graph on the grid in the response sheet.**

```{r fig-grid-temp-day-hour,echo=FALSE,fig.cap="Grid for drawing $f(T)$.",fig.width=4,fig.height=4,fig.align='center',message = FALSE,warning=FALSE}


ggplot() +
  geom_blank()+
  ylab("f(T)") + xlab("h") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) + scale_x_continuous(breaks=seq(0,24,by=4),
                         limits = c(0,25),
                         minor_breaks = seq(0,24,by=2)) +
    ylim(c(0,2))

```

14. Using your graph, explain how the plot of $T(h)$ in @fig-grid-temp-day-hour could be representative of soil temperature over the course of a warm day. **Please answer the question on the response sheet.**

15. Since $f(T)$ ultimately affects $D_{s}$, which in turn is directly proportional to $F_{s}$, interpret your previous graph to determine at what hours of the day would you expect the flux $F_{s}$ to be the largest. At what hours of the day is $F_{s}$ the smallest? Increasing or decreasing the quickest? **Please answer the question on the response sheet.**

### Summarizing up


16. In @tbl-summary indicate if the following factor increases or decreases soil CO$_{2}$ by making a check ($\checkmark$) for yes or $X$ for noin the appropriate column. **Complete the table on the response sheet.**

| Factor        | Increase soil CO$_{2}$ | Decrease soil CO$_{2}$ |
|---------------|------------------------|------------------------|
| Temperature   |                        |                        |
| Soil moisture |                        |                        |
| Plant roots   |                        |                        |
| Soil microbes |                        |                        |

: Evaluation of different factors that influence soil CO$_{2}$. {#tbl-summary}

## Understanding and estimating concentration gradients


17. The two depths closest to the soil surface occur at 3 and 9 cm. Using the values in @tbl-profile1, Compute the rate of change in CO$_{2}$ for each unit change in depth using the two depths closest to the soil surface (this occurs at 3 and 9 cm).  **Please answer the question on the response sheet.**

```{=tex}
\begin{equation}
\frac{dC}{dz} \approx \frac{C_{\mbox{at 9 cm}} - C_{\mbox{at 3 cm}}}{9-3}
\end{equation}
```

| Remember to include units with your answer.


| **Depth (cm)** | **CO**$_{2}$ (ppm) |
|----------------|--------------------|
| 3              | 735                |
| 9              | 1345               |
| 21             | 2879               |

: Data for a CO$_{2}$ profile at different soil depths. {#tbl-profile1}

18. @tbl-profile2 provides data for another CO$_{2}$ profile. Using the information in the table, estimate $\displaystyle \frac{dC}{dz}$.  **Please answer the question on the response sheet.**


| **Depth (cm)** | **CO**$_{2}$ (ppm) |
|----------------|--------------------|
| 3              | 735                |
| 9              | 1345               |
| 21             | 2879               |

: Another CO$_{2}$ profile at different depths. {#tbl-profile2}


19. Which CO$_{2}$ profile (@tbl-profile1 or @tbl-profile2) would predict a larger value of $F_{s}$? Provide a quick explanation why. **Please answer the question on the response sheet.**



## Evaluating patterns in soil fluxes and environmental measurements



20. Study the above interactive figure. Working with your group, identify:

 - time intervals where $F_{s}$ increases rapidly.
 - time intervals where $F_{s}$ steadily increases.
 - time intervals where $F_{s}$ has several fluctuations.
 
| **Please answer the question on the response sheet.**

21. Next co-locate these intervals with measurements of $T$ (@fig-soil-t) and $W$ (@fig-soil-water). Explain how the measurements of $T$ and $W$ during these intervals would lead to larger or smaller values of $F_{s}$. **Please answer the question on the response sheet.**



22. Choose two particular half-hours that correspond to where $F_{s}$ is low and another where $F_{s}$ is high and record your values of $F_{s}$, $T$, and $W$ in @tbl-flux-compare. Compute $f(T)$ and $g(W)$ at each of these times. Explain how these computations provide evidence / support for the calculated value of $F_{s}$. **Please fill in the table on the response sheet.**

| h     | $T$ | $W$ | $f(T)$ | $g(W)$ | $F_{s}$ |     |
|---------|-----|-----|--------|--------|---------|-----|
| $F_{s}$ low:  |     |     |        |        |         |     |
| $F_{s}$ high: |     |     |        |        |         |     |

: Summary table comparing two fluxes {#tbl-flux-compare}

23. Here is a link to nearby weather station located by the San Joaquin Experimental Range: [LINK](https://www.timeanddate.com/weather/@5359383). Look at the historical weather during the dates and intervals you selected to see if you can infer any additional evidence and support for what observed in $T$, $W$, $D_{s}$, or $F_{s}$. **Please answer the question on the response sheet.**

## Additional questions for exploration

24. Which CO$_{2}$ profile (@tbl-profile1 or @tbl-profile2) might produce a higher uncertainty for $\displaystyle \frac{dC}{dz}$? Explain your reasoning. **Please answer the question on the response sheet.**

&nbsp;

25. Draw a conceptual picture of a soil CO$_{2}$ profile following a heavy rainstorm that re-charged the soil, but the day following the top soil layers dried out. Explain why the soil profile should look that way. **Please answer the question on the response sheet.**


